\subsection{Numbers, Fractions, Decimals, Approximations}
	
\begin{enumerate}


			\subsubsection{Evaluation / BODMAS}
%evaluate (BODMAS)
	\item Evaluate $[9876 - 4321] \div 55 - 7 \times 6 + 3$
	
	\item Evaluate $24 \times (10 + 54) \div 8 - 2$
	
	\item Using mathematical tables, evaluate $\cfrac{(36.12)^3 \times 750.9}{(113.2)^2 \times \sqrt{92.5}}$
	
%fraction form	 	
	 \item Evaluate $\cfrac{ \frac{1}{25} + 0.28 + 1\frac{17}{25} }{ 3\frac{1}{5} \div \frac{0.32}{0.15} }$ Give your answer in fraction form.
	 
	 \item Evaluate $\cfrac{0.0084 \times 1.23 \times 3.5}{2.87 \times 0.056}$ with using mathematical tables and express the answer as a fraction in its simplest form.

%numbers - standard form
	\item Given $x = 4.5 \times 10^{-7}$ and $z = 7.2 \times 10^5$, find $y$ in standard form if $z = xy$.
	
	\item Given $x = 1.6 \times 10^8$ and $y = 5.6 \times 10^4$, find $z$ in standard form if $ xz = y$.
	
	\item Compute $0.678 \times 145$ and express the answer in standard form.
	
	\item Given that $M = 4 \times 10^3$, express $\frac{1}{M}$ in standard form.


			\subsubsection{GCF, LCM}
%GCF, LCM	
	\item Find the product of the LCM and GCF of 40, 120 and 240.
	
	\item Find (a) GCF and (b) LCM of 15, 35 and 56.
	
	\item Two numbers, 60 and $n$, have the lowest common multiple (LCM) of 420. If $n$ is a multiple of 6 less than 90, find:
		\begin{itemize}
		\item[(i)] the possible values of $n$.
		\item[(ii)] the greatest common factor (GCF) of the two numbers.
		\end{itemize}	
		
	\item The GCF and LCM of $y$, 18 and 60 are 6 and 360 respectively. Find the value of $y$.
		
	\item The numbers 28, 41, 42, 59, 70 belong to the set of natural numbers. By using these numbers:
		\begin{itemize}
		\item[(i)] calculate the difference between the least common multiple (LCM) of the prime numbers and the greatest common factor (GCF) of the remaining numbers.
		\item[(ii)] express the answer obtained in (i) above in the standard form $A \times 10^n$ correct to two significant figures where $1 \leq A < 10$ and $n$ is an integer.
		\end{itemize}			 
	
	
			\subsubsection{Integers}
%integers
	\item By using properties of integers, give reasons why each of the following expressions are equal:
		\begin{itemize}
		\item[(i)] $(15 \times 17) \times 12 = 17 \times (15 \times 12)$
		\item[(ii)] $15 \times (17 + 12) = (15 \times 17) + (15 \times 12)$
		\end{itemize}
		
	\item An aeroplane traveling at an altitude of 3200 metres makes a climb of 1500 metres, followed by a drop of 100 metres.
		\begin{itemize}
		\item[(a)] Represent the plane's altitude as a sum of positive and negative integers.
		\item[(b)] What is the planes altitude after the drop?
		\end{itemize}
	

		\subsubsection{Recurring Decimals}	 
%recurring decimals
	 
	 \item Express $1.8\dot{6}$ as an improper fraction in its simplest form.
	 
	 \item Change $0.\dot{1}2\dot{3}$ into a fraction.
	 
	 \item Express $23.\dot{1}2\dot{3}$ as a fraction.
	 
	 \item Express $2.\dot{1}35\dot{3}$ as a fraction.
	 
	 \item Express $2.\dot{1}4\dot{6}$ as a fraction.
	 
	 \item Change $0.0\dot{1}$ into a fraction.
	 
	 \item Determine the fractional notation for $0.6\dot{3}$.
	 
	 \item Express $1.\dot{2}1\dot{3}$ as a rational number.
	
	\item Express the following recurring decimal number as a rational number: $0.\dot{6}5\dot{7}$.
	 
	 \item Express $0.8\dot{3}$ in the form $\frac{a}{b}$ where $a$ and $b$ are integers such that $b \neq 0$.
	 
	 \item Express $0.\dot{3}0\dot{3}$ as a fraction, i.e. $\frac{a}{b}$ where $a$ and $b$ are integers and $b \neq 0$.

			\subsubsection{Word Problems}
	\item Mr. Bean lived a quarter of his life as a child, a fifth as a teenager and a third as an adult. He then spent 13 years in his old age. How old was he when he died?	
			
	\item Mary received a certain amount of money from her father to go to school. She spent one third in her journey to school. At school she paid two thirds of the remaining amount as school fees and remained with 24,000 /= as her pocket money. Showing the procedure, calculate the total amount of money she received from her father. 
			
	\item A certain worker used his salary as follows: 20\% on house rent, 45\% on food, 10\% on refreshment and 15\% on school fees. If he\slash she was left with Tsh. 22,000, determine:
		\begin{itemize}
		\item[(i)] the salary of the worker.
		\item[(ii)] the amount of money which he\slash she spent on food.
		\end{itemize}

		\subsubsection{Approximations}	
%approximations - give to _ decimal places
	\item By using mathematical tables, evaluate $\cfrac{\sqrt[3]{0.0072} \times (81.3)^2}{\sqrt{23140}}$ to three significant figures.
	
	\item By using mathematical tables, evaluate $\cfrac{(28.32)^4 \times 0.03574}{87.57}$. Give your final answer in standard form, to three significant figures.
	
	\item If $a = 2.432 \times 10^4$, $b = 7.42 \times 10^{-2}$ and $c = 0.0324 \times 10^{-2}$, find the value of $R$ in standard form correct to two decimal places given that $R = \cfrac{ab}{c}$.
	
	\item Using mathematical tables, evaluate the expression: $\cfrac{(28.32)^2 \times 0.3574}{\sqrt{8.732}}$. Give the answer correct to four decimal places.
	 
	 \item By using mathematical tables evaluate the expression $\cfrac{237.8 \times 0.0873}{67890}$. Give the answer to four figures.
	 
	 \item Given that $\sqrt{3} = 1.7321$, calculate the value of $\cfrac{2}{\sqrt{3} - 1}$ correct to 4 decimal places.

%round off, estimate
	\item Estimate he value of $\cfrac{57.2 \times 110}{2.146 \times 46.9}$ correct to one (1) significant figure.
	 
	 \item 
	 	\begin{itemize}
	 	\item[(a)] Find the value of the expression $\left(\cfrac{4.75 + 1.31}{3.13}\right)^2$ giving your answer in three decimal places.
	 	\item[(b)] By rounding each term of the expression in (a) above to one significant figure, obtain a rough estimate of the expression.
	 	\end{itemize}
	 
	 \item 
	 	\begin{itemize}
	 	\item[(a)] Round off each of the following numbers to one decimal place. $L = 20.354$, $M = 40.842$, $N = 10.789$
	 	\item[(b)] Use the result obtained in (a) above to find the value of $x$ given that $x = \cfrac{LM}{N}$.
	 	\end{itemize}
	 	
	\item Round each of the numbers $x = 2.354$, $y = 4.843$ and $z = 1.789$ to one decimal place and then use the results obtained to find the value of $A$ to two significant figures given that $A = \cfrac{xy}{z}$.
	 
	\item Express 0.05473
		\begin{itemize}
		\item[(i)] correct to three (3) significant figures
		\item[(ii)] correct to three (3) decimal places
		\item[(iii)] in standard form
		\end{itemize}
	
	\item Write 624.3278 correct to
		\begin{itemize}
		\item[(i)] five significant figures
		\item[(ii)] three decimal places
		\end{itemize}



\end{enumerate}	
